Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-09 docs/Manuscript/Manuscript.Rmd

We discuss in this review a hybrid computational approach we call _localized modeling_ that appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which may deviate from the index case.
A prediction or inference is then obtained from a model fitted to this _similarity cohort_.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

@@

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort can be understood as an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a ML workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to predict classes or outcomes. Nevertheless, as some recent applications have begun to show, they can also be used to draw individualized inferences about the importance of different risk factors.
While most ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions, these quantifications are not directly interpretable model components analogous to the split nodes of a decision tree (DT) or the effect estimates of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model as a whole inherits this property. Thus, localized models may embody both styles of CBR.

@@

Our methods followed prepared plans, which we describe in detail in the Appendix.
**Additional details are provided in Section \label{sec:appendix-selection}.**
We also note deviations from these plans and the reasons for them.

@@

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators (Kohonen + CBR [@Wyns2004a], TWNFI [@Song2006c, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015a], CBR + rules (+ DML) [@Nicolas2014a], Gaussian process regression [@Zhang2018a], JITL-ELM [@Ma2020a]) are hybrids of localized modeling (in some cases CBR) with other techniques, often metric learning. Though @CampilloGimenez2012a and [@Ng2015] report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model (statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]), they did not consistently outperform comparators.

@@

We draw several observations from our encoding of the included studies. We first focus on two driving questions: What challenges did investigators set out to solve? How did their approaches meet these challenges? We then comment on the connectedness of the included studies and the potential for recombination.

@@

## Coherence

@@

The 25 studies included in our synthesis used composite techniques that we organized into two major types. One type, similarity learning, was used in 8 studies that tie in to the much larger literature on metric learning [@Yang2006]. These techniques are designed to detect structure in data, especially associations between predictors and responses, and to use this structure to inform the definition of a patient similarity measure. Seven studies used response values in training data sets to supervise similarity learning while two used unsupervised learning (Table \ref{tab:composite}). Based on the key properties of metric learning algorithms identified by @Bellet2014, most learned measures were non-linear while some were local, and some were optimized globally while others locally. However, not all qualified as metric learning, since the resulting measures would not necessarily satisfy the triangle inequality. Each then used its similarity measure to retrieve training cases relevant to each testing case. Once retrieved, most took a standard NN approach to generating predictions from the cohort; two exceptions [@Vilhena2016a; @Tang2021c] proceeded to fit predictive models to the retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several studies used unsupervised similarity learning, for example the Mahalanobis distance [@Lowsky2013] and the Kolmogorov entropy-based distance [@Elter2007a]. In particular, @Yearwood1997a defined similarity as a weighted sum of differences in predictor values and used linear regression to optimize the weights for the predictive accuracy of the cohort they retrieve. Others used supervised learning: @Song2006c proposed an iterative algorithm to optimize a set of fuzzy inference rules used to retrieve relevant cases and generate a prediction, using back-propagation on the rules’ parameters. Their method was used in several later studies returned by our search, which are not discussed here because they did not originate the technique. @Nicolas2014a explicitly appeal to a supervised distance metric learning (DML) technique [@Xing2002] to obtain a similarity measure on their set of dermoscopy and confocal images that most effectively separates malignant from benign melanoma tumors. We will discuss the remaining uses of supervised similarity learning shortly.

@@

As noted above, these instances of similarity learning fit into a much larger literature on metric learning, which has seen widespread use in health informatics and modeling. That these studies satisfied our review criteria is due to the novelty of the techniques at the time of publication and the detailed attention paid to their techniques by the authors. It also reflects a limitation of our study and of the literature we set out to retrieve: We are aware of no standard terms in use to identify the family of techniques that fit statistical models to similarity-based cohorts. The retrieved studies whose use of such techniques alone satisfied our criteria variably termed them individualized, personalized, local, and patient-specific models, among other terms (Table \ref{tab:composite}). We prefer the term localized models, which is both sensitive and specific to uses of a similarity or distance measure to retrieve a cohort to which a model is fit: It includes some properly included techniques we did not have in mind [@Vilhena2016a] yet excludes other techniques commonly referred to as individualized, personalized, or precision.

@@

This relatively small sample exhibited a wide range of approaches, both technically and conceptually. Discussion of the technical diversity of these approaches has been summarized above and published in greater detail in previous reviews [@Choudhury2016; @Sharafoddini2017; @Parimbelli2018], and we focus here on the conceptual. The design of a localized modeling approach can be decomposed into three largely independent choices: (1) how to measure the similarity between cases, (2) how to retrieve a cohort of cases similar to an index case, and (3) how to generate a prediction or other statistical insight for the index case from the similarity cohort. The choices are the _similarity measure_, the _retrieval step_, and the _statistical model_, respectively. These are diagrammed in Figure \ref{fig:framework}.

@@

Each step can be unsupervised supervised: Unsupervised similarity measures include discrete measures like the Levenshtein distance and continuous measures like cosine similarity, while supervised measures include the Mahalanobis distance and random forest proximity as well as several composite measures with weights calculated from the data. Most retrieval steps obtained cohorts with a uniform size (cardinality) or similarity threshold, and the remaining were likewise only informed by predictors; thus, while a retrieval step could in principle be supervised, in this sample none were. Almost all models were supervised, being that most performed predictive tasks, though @Mariuzzi1997 modeled only survival rates within localized cohorts. Meanwhile, each step can be optimized in a ML fashion by having its parameters tuned to improve performance. We found no studies that tuned the calculation of similarity, though most uniform cohort sizes were tuned. Some earlier studies tuned generalized regression models, but more recent studies have not.

>>>>

We discuss in this review a hybrid computational approach we call _localized modeling_ (see Section \ref{sec:coherence}) that appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which may deviate from the index case.
A prediction or inference is then obtained from a model fitted to this _similarity cohort_.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

@@

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort can be understood as an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a ML workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to predict classes or outcomes.
Nevertheless, as some recent applications have begun to show, they can be used to draw inferences about the importance of different risk factors to specific individuals.
While most ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions, these quantifications are not directly interpretable model components analogous to the split nodes of a decision tree (DT) or the effect estimates of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model as a whole inherits this property. Thus, localized models may embody both styles of CBR.

@@

Here we describe our review process, including deviations from plans and the reasons for them.
More details are included in Section \label{sec:appendix-selection}.

@@

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators (Kohonen + CBR [@Wyns2004a], TWNFI [@Song2006c, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015a], CBR + rules (+ DML) [@Nicolas2014a], Gaussian process regression [@Zhang2018a], JITL-ELM [@Ma2020a]) are hybrids of localized modeling (in some cases CBR) with other techniques, often DML. Though @CampilloGimenez2012a and [@Ng2015] report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model (statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]), they did not consistently outperform comparators.

@@

After commenting briefly on our review process, we draw several observations from our encoding of the included studies.
We first focus on their methodological choices and innovations, then on their conventions and language.
We spend most of the section discussing the differences and commonalities of the techniques used, toward a general method that encompasses most or all cases.
We then suggest some avenues for future studies and conclude with our key takeaways.

@@

## Coherence
\label{sec:coherence}

@@

The 25 studies included in our synthesis used composite techniques that we organized into two major types. One type, _similarity learning_, was used in 8 studies that tie in to the much larger literature on DML [@Yang2006].
The other, _cohort thresholding_, focused on choosing or optimizing the manner in which cohorts were demarcated using the similarity measure.

@@

The similarity learning approaches took a variety of forms.
These techniques are designed to detect structure in data, especially associations between predictors and responses, and to use this structure to inform the definition of a patient similarity measure. Seven studies used response values in training data sets to supervise similarity learning while two used unsupervised learning (Table \ref{tab:composite}). Based on the key properties of DML algorithms identified by @Bellet2014, most learned measures were non-linear while some were locally learned, and some were optimized globally while others locally. Not all qualified as DML, since the resulting measures would not necessarily satisfy the triangle inequality. Each used its similarity measure to retrieve training cases relevant to each testing case. Once retrieved, most took a standard NN approach to generating predictions; two exceptions [@Vilhena2016a; @Tang2021c] fit predictive models to the retrieved cohorts.

Several studies used unsupervised similarity learning, for example the Mahalanobis distance [@Lowsky2013] and the Kolmogorov entropy-based distance [@Elter2007a]. In particular, @Yearwood1997a defined similarity as a weighted sum of differences in predictor values and used linear regression to optimize the weights for the predictive accuracy of the cohort they retrieve. Others used supervised learning: @Song2006c proposed an iterative algorithm to optimize a set of fuzzy inference rules used to retrieve relevant cases and generate a prediction, using back-propagation on the rules’ parameters. Their method was used in several later studies returned by our search, which are not discussed here because they did not originate the technique. @Nicolas2014a explicitly appeal to a supervised distance metric learning (DML) technique [@Xing2002] to obtain a similarity measure on their set of dermoscopy and confocal images that most effectively separates malignant from benign melanoma tumors. We will discuss the remaining uses of supervised similarity learning shortly.

@@

As noted above, these instances of similarity learning fit into a much larger literature on DML, which has seen widespread use in health informatics and modeling. That these studies satisfied our review criteria is due to the novelty of the techniques at the time of publication and the detailed attention paid to their techniques by the authors. It also reflects a limitation of our study and of the literature we set out to retrieve: We are aware of no standard terms in use to identify the family of techniques that fit statistical models to similarity-based cohorts. The retrieved studies whose use of such techniques alone satisfied our criteria variably termed them individualized, personalized, local, and patient-specific models, among other terms (Table \ref{tab:composite}). We prefer the term localized models, which is both sensitive and specific to uses of a similarity or distance measure to retrieve a cohort to which a model is fit: It includes some properly included techniques we did not have in mind [@Vilhena2016a] yet excludes other techniques commonly referred to as individualized, personalized, or precision.

@@

This relatively small sample exhibited a wide range of approaches, both technically and conceptually. Discussion of the technical diversity of these approaches has been summarized above and published in greater detail in previous reviews [@Choudhury2016; @Sharafoddini2017; @Parimbelli2018], and we focus here on the conceptual. The design of a localized modeling approach can be decomposed into three largely independent choices: (1) how to measure the similarity between cases, (2) how to retrieve a cohort of cases similar to an index case, and (3) how to generate a prediction or other statistical insight for the index case from the similarity cohort. The choices are the _similarity measure_, the _cohort retrieval_, and the _statistical model_, respectively. These are diagrammed in Figure \ref{fig:framework}.

@@

Each step can be unsupervised supervised: Unsupervised similarity measures include discrete measures like the Levenshtein distance and continuous measures like cosine similarity, while supervised measures include the Mahalanobis distance and random forest proximity as well as several composite measures with weights calculated from the data. Most retrieval steps obtained cohorts with a uniform size (cardinality) or similarity threshold, and the remaining were likewise only informed by predictors; thus, while a retrieval step could in principle be supervised, in this sample none were. Almost all models were supervised, being that most performed predictive tasks, though @Mariuzzi1997 modeled only survival rates within localized cohorts. Meanwhile, each step can be optimized in a ML fashion by having its parameters tuned to improve performance. We found no studies that tuned the calculation of similarity, though most uniform cohort sizes were tuned. Some earlier studies tuned generalized regression models, but more recent studies did not.
